---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# famr

<!-- badges: start -->
<!-- badges: end -->

The goal of famr is to fit a Bayesian factor model for multivariate, mixed responses that addresses multicollinearity in the predictors and provides more precise estimates through the sharing of information between responses.

## Installation


You may install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("phuchonguyen/famr")
```

## Example

This is a basic example which shows you how to use FAMR to model two outcomes given six highly correlated predictors and one covariate:

```{r loadlib}
library(famr)
```

```{r makedat, include=FALSE}
set.seed(111)
n <- 100
K <- 2
p <- 6
t <- 3
Eta <- matrix(rnorm(n*K), n, K)
colnames(Eta) <- paste0("e", 1:K)
Theta <- matrix(c(rep(rep(c(1,0), each=p/2), K/2), rep(rep(c(0,1), each=p/2), K/2)), p, K) # block structure
Theta <- abs(rnorm(K*p, 2))*Theta
sigmax <- rep(.5, p)
sigmax_sqinv <- 1/sigmax^2
X <- tcrossprod(Eta, Theta) + t(matrix(rnorm(n*p, rep(0, p), sigmax), p, n))
Z_int <- array(rnorm(n), dim=c(n, 1))
Eta_int <- model.matrix(~. - 1, data = as.data.frame(Eta))
Eta_int <- cbind(Eta_int, apply(Eta, 2, function(x) x*Z_int))
K_int <- ncol(Eta_int)
B <- matrix(c(.75, 0, 0, 0, 
              2.0, 0, 0, 0,
              1.2, 0, 0, 0), 4, 3)
maineffects <- diag(sigmax_sqinv) %*% Theta %*% solve(t(Theta) %*% diag(sigmax_sqinv) %*% Theta + diag(1, K)) %*% B[1:2, ]
Eta_extra <- matrix(rnorm(n*K), n, K)
B_extra <- matrix(rnorm(t*K, rep(0, K), 1), K, t)
sigmay <- diag(runif(t, 0.7, 1.5))
Sigma <- crossprod(B_extra, B_extra) + sigmay
Y1 <- Eta_int%*%B + Eta_extra%*%B_extra +  MASS::mvrnorm(nrow(Eta), rep(0, t), sigmay)
Y2 <- Eta_int%*%B + Eta_extra%*%B_extra +  MASS::mvrnorm(nrow(Eta), rep(0, t), sigmay)
Y <- rbind(Y1, Y2)
id <- c(1:nrow(Y1), 1:nrow(Y2))
X=rbind(X, X)
Z_int=rbind(Z_int, Z_int)
```


Fit the model assuming the six correlated predictors come from two latent factor

```{r fitmod}
s <- Sys.time()
out <- famr(Y=Y, X=X, K=2, Z_int=Z_int, niter=5000, id=id, varimax=TRUE)
print(Sys.time() - s)
```

Check trace plots for convergence

```{r diagnostics, echo=FALSE}
par(mfrow=c(1,3))
for (i in 1:3) {
  matplot(out$Bx[,,i], type="l")
}
```


Compare the inferred effects with the truth

```{r result, echo=FALSE}
par(mfrow=c(1,2))
image(B, main="true effects")
image(apply(out$B_varimax[,,], c(2,3), mean), main="posterior mean with varimax")
```


